---
title: 生物统计原理
author: "王强"
institute: "南京大学生命科学学院"
date: \today{}
toc: true

---

```{r setup, include=FALSE}
# load packages
library(knitr)
library(tidyverse)
library(lubridate)
library(survival)
library(survminer)
library(randomNames)

# set output options
knitr::opts_chunk$set(fig.width = 4, fig.height = 4, dpi=300)
options(digits = 4)
```

```{text compile, include=FALSE}
# Rscript -e "library(knitr); knitr::knit('biostat-survival.rmd');"
# make biostat-survival

# Rscript -e "library(randomNames); randomNames(10, ethnicity=5, which.names='first', sample.with.replacement=F);"

```

# 生存分析

## 定义

* 生存分析
: 是统计的一个分支，用于分析一个或多个事件发生前的预期持续时间.

* Survival analysis

* 事件 - 例如生物体死亡和机械系统故障

\note{

生存分析试图回答以下问题：

将在一定时间后生存的人口比例是多少？

在那些幸存者中，他们将以何种速度死亡或失败？

可以考虑多种死亡或失败原因吗？

特定情况或特征如何增加或降低生存概率？

事件可以是非常明确的, 例如死亡; 但也可能有些模糊, 心肌梗死, 器官衰竭等

对于机械系统, 故障可能无法明确定义

}

---

* 工程学 - 可靠性分析
* 经济学 - 持续时间分析
* 社会学 - 事件历史分析

## 数据

对于癌症

* Overall survival (OS)
  * Time from surgery to death
* Progression-free survival (PFS)
  * Time from start of treatment to progression, Time to progression (TTP)
  * Time from response to recurrence

## 分析方法

* 描述生存时间

  * 生命表
  * Kaplan--Meier 曲线
  * 生存函数
  * 风险函数

* 比较两组或多组生存时间

  * 秩和检验

* 变量对生存的影响

  * Cox 比例风险回归

\note{

保险公司里精算师 生命表（也称死亡率表或精算表）

}

# Kaplan--Meier curve

## Basic Kaplan--Meier curve

```{r basic, warning=F, echo=F, message=F, comment=NA}
followup <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
eventtype <- c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1)

patientID <- c("Sarah", "Joel", "James", "Daniel", "Eric", "Kaitlynn", "Kelly", "Morgen", "Brandie", "Luke" )
scenario <- c("A", "A", "A", "A", "A", "A", "A", "A", "A", "A")

toy_dataset <- data.frame(patientID, followup, eventtype, scenario)
colnames(toy_dataset) <- c("PatientID", "FollowUp", "EventType", "Scenario")

kable(toy_dataset)

```

\note{

从创建一些简单的数据开始

10 个高加索人种的病人, 10 个月的随访数据, 对所有参与者, 同样的治疗手段

If we take a closer look at the 'Follow-up' and 'Eventtype' columns:

* Event type of 1 equals an event (for example death in an overall survival study)
* Event type of 0 equals a right-censored event.

In this first example, there are no censored events.

}

---

随访
: Follow up, 可以用任何时间间隔, 天、月、年

事件
: Event type,　`1` 表示事件发生, 在 Overall Survival 里, 代表死亡

分层
: Strata, Scenario

拟合
: Fit

\note{

Strata

英  ˈstrɑːtə   美  ˈstreɪtə

男/女, 给药/安慰剂, 高/低分险 等

Note that in this simulation we are showing multiple scenarios, so we are fitting to
`Scenario`, but this can also be for example a treatment column.

}

---

```{.r .numberLines}
# The R package `survival`
library(survival)

# Define what the time column is toy_dataset$FollowUp
#   and the event column is toy_dataset$EventType
surv_object <- Surv(
    time = toy_dataset$FollowUp,
    event = toy_dataset$EventType
)

# Fit the Kaplan--Meier curve to this data `Scenario`
fit <- survfit(
    surv_object ~ Scenario,
    data = toy_dataset
)
```

---

```{r km, fig.cap = 'KM - basic', warning=F, echo=F, message=F}
surv_object <- Surv(time = toy_dataset$FollowUp , event = toy_dataset$EventType)
fit1 <- survfit(surv_object ~ Scenario, data = toy_dataset)
j <- ggsurvplot(fit1,
    data = toy_dataset,
    conf.int = FALSE,
    risk.table = "nrisk_cumcensor",
    break.time.by = 2,
    risk.table.fontsize = 3,
    risk.table.height = 0.20,
    censor.shape = "|",
    censor.size = 3,
    ggtheme = theme_bw()
)

j$plot <- j$plot +
    theme(aspect.ratio=1) +
    theme(panel.grid.major.x = element_blank(), panel.grid.major.y = element_blank()) +
    theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank()) +
    scale_x_continuous(limits = c(0,10), breaks = sort(c(seq(0, 10, 1)))) +
    scale_y_continuous(limits = c(0,1), breaks = seq(0, 1, 0.2), expand = c(0.02,0.02)) +
    theme(legend.background = element_rect(colour = "transparent", fill = "transparent")) +
    theme(legend.justification=c(1,1), legend.position=c(1,1), legend.direction = "vertical") +
    theme(legend.title = element_blank())

j$table <- j$table +
    theme_cleantable() +
    theme(plot.title = element_text(size=8))

j

```

\note{

Every month, 1 participant has an event (in an overall survival study this would be death).

Every time an event occurs, the survival probability drops by 0.1 of the remaining curve
(= events / at risk = 1/10 of the remaining curve at month 1,
1/9 of the remaining curve at month 2 and so on)

}

## Kaplan--Meier curve with censored data

```{r censor, warning=F, echo=F, message=F, comment=NA}
followup <- c(1, 2, 3, 4, 5, 6, 6.2, 8, 9, 10)
eventtype <- c(1, 1,  0, 0, 1, 1, 1, 0, 1, 0)

toy_dataset <- data.frame(patientID, followup, eventtype, scenario)
colnames(toy_dataset) <- c("PatientID", "FollowUp", "EventType", "Scenario")

kable(toy_dataset)

```

---

删失
: Censored, Censoring

---

```{r km_censored, fig.cap = 'KM - censored', ref.label='km', warning=F, echo=F, message=F}
```

## Importance of confidence intervals

```{r scenario, warning=F, echo=F, message=F, comment=NA}
followup <- c(1, 2, 3, 4, 5, 6, 6.2, 8, 9, 10, 1, 2, 3, 4, 5, 6, 6.2, 8, 9, 10)
eventtype <- c(1, 1, 0, 0, 1, 1, 1, 0, 1, 0, 1, 1, 0, 0, 1, 1, 1, 0, 0, 1)

scenario <- rep(c("A", "B"), each = 10)

toy_dataset <- data.frame(patientID, followup, eventtype, scenario)
colnames(toy_dataset) <- c("PatientID", "FollowUp", "EventType", "Scenario")

for (value in unique(toy_dataset$Scenario)){
    print(tail((toy_dataset[toy_dataset$Scenario == value,]),2))
}

# small offset otherwise curves would almost fully overlap.
toy_dataset$FollowUp = ifelse(toy_dataset$Scenario == "A", toy_dataset$FollowUp + 0.1, toy_dataset$FollowUp + 0)

```

\note{

Especially when there are very few patients at risk, the impact of a censored event can have a big
impact on the appearance of the KM curve.

In the previous plot, it seems that the survival curve reaches a plateau at 0.2 surival probability
(curve A). If we would swap the censored status between Joe and Kate (participants 9 and 10), the KM
curve changes drastically and drops to 0 at the end of the study period. In this scenario (curve B),
all participants either had an event or were censored.

}

---

```{r km_scenario, fig.cap = 'KM - scenarios', ref.label='km', warning=F, echo=F, message=F}
```

